CODECHECK: An open-science initiative for the independent execution of computations underlying research articles

### 2021-01-12 codecheck-paper.tex
update introduction (clarify, ..)

Many areas of scientific research now use computation to either simulate
or analyse their data. These computations are increasingly complex, and
difficult to explain coherently in a paper \citep{marwick_how_2015}.
To complement the traditional route of sharing research by writing papers,
there is a growing trend/demand to share the underlying artefacts, notably 
code and datasets, so that others can inspect, reproduce or expand that work
(see Figure~\ref{fig:inverse}).
Some of the earliest proponents were Buckheit and Donoho
\cite{buckheit_wavelab_1995} who coined what has been called 
\emph{Claerbout's claim} (extending on \citet{claerbout_electronic_1992}):
\emph{"An article about computational science in a scientific publication 
is \text{not} the scholarship itself, it is merely \textbf{advertising} of
the scholarship. The actual scholarship is the complete software development
environment and the complete set of instructions which generated the 
figures."}
%  Donoho (2010) has another paraphrasing of the same quote, which might work better:
% "an article about computational result is advertising, not scholarship. The actual scholarship is the full software environment, code and data, that produced the result."

So, assuming that researchers begin to share more artefacts, how might
they be examined to test that they do what they claim? For example,
most scientific journals now require a data sharing statement that
outlines what data the authors have (or will) share.
The implementation status of this varies considerably according to the
journal. At one of the spectrum, there are specialist journals that
have been created to accept ``data papers''
(e.g.~\href{https://www.nature.com/sdata/}{\emph{Scientific~Data}}, 
\href{https://essd.copernicus.org/}{\emph{Earth System Science Data}}, 
\href{https://rmets.onlinelibrary.wiley.com/journal/20496060}{\emph{Geoscience Data Journal}},
\href{https://bdj.pensoft.net/}{\emph{Biodiversity Data Journal}},
\href{https://openpsychologydata.metajnl.com/}{\emph{Journal of Open Psychology Data}},
\href{https://odjar.org/}{\emph{Open Data Journal for Agricultural Research}},
\href{https://openhealthdata.metajnl.com}{\emph{Journal of Open Health Data}}).
These journals have established rigorous procedures by which
the data is validated according to standards in each field. At the
other end of the spectrum, we still witness today the infamous
statement ``Data available upon reasonable request''. Authors,
while possible well-intentioned at the time of writing the article,
often cannot provide the data when readers ask for
it as data disappears over time \cite{Vines2014-hf}.

Given that there are no clear standards yet for sharing data, what hope
might there be for sharing computer programs? While both data and software
together are required to validate outcomes of a computational analysis,
they do differ in  that data can be seen as static/inert,
while code requires an environment and interaction. This makes 
software harder to share.
Our anectodal experiences
on this matter suggest that often a variety of reasons are given for why
code cannot be shared, e.g.~``there is no documentation / I do not
want to maintain it / I do not want to give away my code to my
competitors''. Our view, rooted in Open Science principles, is that
wherever possible sharing code is good for the community, as outlined
in 2010 \cite{Barnes2010-iv}. Having the code freely available, and
suitably archived, provides a valuable resource for others to learn
from, even when it doesn't run or if there is no documentation. However,
with a little effort, we believe that if someone independent can run/execute
the code, this is worth documenting as early as possible and that the barrier
to evaluating non-text parts of research can be greatly reduced with a 
focused approach.
Just like peer review can provide a ``baseline reassurance'', i.e., a paper
has at least been checked by someone with an understanding of the topic
\cite{fyfe_mission_2019}, we think the same baseline should be provided
regarding the workflow underlying a paper.
With this in mind, we have developed a set of PRINCIPLES and an example 
workflow that provides what he hope is a pragmatic way of checking that 
code works.

The main contribution of this work is a thorough description of a process
and its variations to integrate a much needed evaluation of the 
computational reproducibility \cite{barba_terminologies_2018} into peer
review, and a demonstration of its feasibility by means of 25
reproductions across different scientific disciplines.
We call this system CODECHECK as outlined next.

>>>>

Many areas of scientific research use computations to either simulate
or analyse their data. These computations are increasingly complex, and
difficult to explain coherently in a paper \citep{marwick_how_2015}.
To complement the traditional route of sharing research by writing papers,
there is a growing demand to share the underlying artefacts, notably 
code and datasets, so that others can inspect, reproduce or expand that work
(see Figure~\ref{fig:inverse}).
Some of the earliest proponents were Buckheit and Donoho
\cite{buckheit_wavelab_1995} who coined what has been called 
\emph{Claerbout's claim} (extending on \citet{claerbout_electronic_1992}):
\emph{"An article about computational science in a scientific publication 
is \text{not} the scholarship itself, it is merely \textbf{advertising} of
the scholarship. The actual scholarship is the complete software development
environment and the complete set of instructions which generated the 
figures."}
%  Donoho (2010) has another paraphrasing of the same quote, which might work better:
% "an article about computational result is advertising, not scholarship. The actual scholarship is the full software environment, code and data, that produced the result."

So, assuming that researchers begin to share more artefacts, how might
they be examined to test that they do what they claim? For example,
most scientific journals now require a data sharing statement that
outlines what data the authors have (or will) share.
The implementation status of this varies considerably according to the
journal. At one end of the spectrum, there are specialist journals that
have been created to accept ``data papers''
(e.g.~\href{https://www.nature.com/sdata/}{\emph{Scientific~Data}}, 
\href{https://essd.copernicus.org/}{\emph{Earth System Science Data}}, 
\href{https://rmets.onlinelibrary.wiley.com/journal/20496060}{\emph{Geoscience Data Journal}},
\href{https://bdj.pensoft.net/}{\emph{Biodiversity Data Journal}},
\href{https://openpsychologydata.metajnl.com/}{\emph{Journal of Open Psychology Data}},
\href{https://odjar.org/}{\emph{Open Data Journal for Agricultural Research}},
\href{https://openhealthdata.metajnl.com}{\emph{Journal of Open Health Data}}).
These journals have established rigorous procedures by which
the data is validated according to standards in each field. At the
other end of the spectrum, we still witness today the infamous
statement ``Data available upon reasonable request''. Authors,
while possible well-intentioned at the time of writing the article,
often cannot provide the data when readers ask for
it as data disappears over time \cite{Vines2014-hf}.

Given that there are no clear standards yet for sharing data, what hope
might there be for sharing computer programs? While both data and software
together are required to validate outcomes of a computational analysis,
they do differ in  that data can be seen as static and inert,
while code requires an environment to be run in and interacted with.
This makes software harder to share.
Our anectodal experiences
on this matter suggest that often a variety of reasons are given for why
code cannot be shared, e.g.,~``there is no documentation'', ``I cannot 
maintain it'', or ``I do not want to give away my code to competitors''.
Our view, rooted in Open~Science principles
\footnote{\url{https://en.wikipedia.org/wiki/Open_science}}, is that
wherever possible sharing code is good for the community and the 
individual \cite{Barnes2010-iv,markowetz_five_2015}.
Having code and data openly available and archived, 
provides a valuable resource for others to learn
from, even when it doesn't run anymore or if there is no documentation.
However,
with a little effort, we believe that if someone independent can execute
an analysis, this is worth documenting as early as possible and that the barrier
to evaluating non-text materials of research can be greatly reduced with a 
focused approach.
Just like data journals' validations of data and all journals' peer review
can provide a ``baseline reassurance'', i.e.,~a paper
has been checked by someone with an understanding of the topic
\cite{fyfe_mission_2019}, we think the same baseline should be provided
regarding the workflow underlying a paper.
With this in mind, we have developed a set of principles and an example 
workflow that provides what he hope is a pragmatic way of checking that a
paper's code works.

The main contribution of this work is a thorough description of a process
and its variations to integrate a much needed evaluation of the 
computational reproducibility\footnote{Following the 
Claerbout/Donoho/Peng terminilogy, see 
\cite{barba_terminologies_2018}} into peer
review, and a demonstration of its feasibility by means of 25
reproductions across different scientific disciplines.
We call this system CODECHECK as outlined next.
